Iterative Monomorphisation

### 2025-02-13 iterative_monomorphisation.tex
edits

Similar bounds are used for the non-monomorphic type arguments tuples:
The limit on the number of new non-monomorphic type argument tuples is \(\min(\max(\PolyMult\cdot n, \PolyFloor),\allowbreak \PolyCap)\), where \(n\) is the number of non-monomorphic type argument tuples associated with the current formula. A significant difference with the monomorphic case is that \(n\) depends on the current formula being processed whilst \(m\) does not.
Both in the monomorphic and in the non-monomorphic case, the maximum number of newly generated type argument tuples is fixed per formula and per iteration.

@@

The monomorphisation algorithm is parametrised by many bounds. The first part of the evaluation process is to find appropriate values for these bounds. The second part compares the performance of Zipperposition with and without E on polymorphic problems. The third part compares the performance of different provers on polymorphic problems and their monomorphised counterparts.

The benchmarks are taken from version 8.2.0 of the TPTP library~\cite{tptp}. The library contains 1765 problems in the TF1 and TH1 grammars, corresponding respectively to first-order and higher-order logic with rank-1 polymorphism. Because Zipperposition does not support reasoning with real numbers, we removed all problems that include them. In total, our benchmark suite contains 1534 polymorphic problems. We chose as a measure of success for a given prover (or prover configuration) the number of problems that could each be solved by the prover in under 30 seconds with a single thread. This corresponds to the default time limit in Sledgehammer.

@@

We conducted additional informal evaluations on the 500 problem benchmark suite to find appropriate default values and test ranges for the monomorphisation bounds. We started the option evaluation process with the base configuration and the following default values for monomorphisation bounds and parameters for E:

\begin{multicols}{2}% 3 columns?
\begin{itemize}
   \item \PolyCap: \(\infty\)
   \item \PolyMult: \(1\)
   \item \PolyFloor: \(50\)
   \item substitution ordering: separation
   \item \SubstLimit: \(\infty\)
   \item \MonoSubstsLimit: \(10\)
   \item \Limit: \(2000\)
   \item new formulae limit multiplier: \(0\)
   \item monomorphisation timeout: \(20\)
   \item \Loop: \(4\)
   \item E timeout: \(30\)
   \item E call point: \(0\)
\end{itemize}
\end{multicols}
\noindent
The initial values for the \MonoCap, \MonoMult and \MonoFloor options are irrelevant because the options' values are set when computing Table~\ref{mono_ty_args}.% and used for subsequent tables.

@@

Table~\ref{nmon_ty_args} plays the same role as Table~\ref{mono_ty_args} except that it evaluates the bounds limiting the number of new \emph{non-}monomorphic type arguments. The bound values are lower because we found non-monomorphic type arguments to be explosive in preliminary evaluations. The table confirms that non-monomorphic type argument generation drives the creation of useful non-monomorphic formulae. This is indicated by the very low number of problems solved when no new non-monomorphic type arguments are allowed. Performance of the monomorphisation algorithm seems to plateau for certain ranges of the tested values and drops off beyond.
%TODO check that 125 is what we get if we just run Zipperposition for 10 seconds (maybe with all additional problems that are easily monomorphised)

@@

%The substitution generation phase occurs once all type arguments have been generated. The bounds limiting the number of new substitutions and the number of monomorphising substitutions per formula are directly related through the role they play in the construction of monomorphising substitutions. The substitution ordering heuristic has a significant impact on monomorphising substitution generation. The ``age'' ordering of substitution orders substitutions generated in earlier iterations are ordered first. The ``random'' ordering randomly shuffles substitutions. Finally, the ``separation'' ordering separates the substitutions into different sets depending on which iteration they were generated in and generates monomorphising substitutions from each of these sets independently. Table \ref{subst_gen_table} shows that the values of bounds limiting the number of substitutions generated has very little impact on overall performance contrary to the ordering heuristic.
The substitution generation phase occurs once all type arguments have been generated. The bound limiting the number of monomorphising substitutions per formula is directly related to the ordering which dictates how such monomorphising substitutions are generated. The substitution ordering heuristic has a significant impact on monomorphising substitution generation. The ``age'' ordering of substitution orders substitutions generated in earlier iterations are ordered first. The ``random'' ordering randomly shuffles substitutions. Finally, the ``separation'' ordering separates the substitutions into groups of substitutions generated in the same iteration and generates monomorphising substitutions from each of these groups independently. Table \ref{subst_gen_table} shows that the values of bounds limiting the number of monomorphising substitutions only seems to affect performance when the ``separation'' heuristic is used.
% TODO find a better term than "separation heuristic" maybe separation ?

@@

For larger problems, the monomorphisation algorithm may time out despite the bounds. In these cases neither Zipperposition nor E will have had a chance to attempt to solve the problem. To avoid this, a timer can interrupt the monomorphisation algorithm, after which Zipperposition resumes normal operation. Table \ref{mono_time} tests the amount of time that monomorphisation is allowed to run for against the number of iteration of the monomorphisation algorithm. Neither parameter seems to have a significant effect on the performance of the monomorphisation algorithm.

@@

\centering\begin{tabular}{@{}l*{3}{>{\centering\arraybackslash}p{6em}}@{}}
   \toprule
   & without E & with E & union \\
   \midrule
   500 problem benchmark   & 160 & 198 & 207 \\ % -9 +47
   1034 problem benchmark & 337 & 410 & 434 \\ % -24 +97
   \bottomrule
\end{tabular}
\label{zipp_eval}
\end{table}

@@

\begin{table}[ht]
\caption{Evaluation of native polymorphism vs.\ monomorphisation}

@@

To evaluate the usefulness of iterative monomorphisation as an alternative to native polymorphism, two competitive higher-order polymorphic provers were run on the 1034 problem benchmark. We chose Leo-III and Zipperposition. We unfortunately had to exclude Vampire because of parsing issues.

>>>>

Similar bounds are used for the non-monomorphic type arguments tuples:
The limit on the number of new non-monomorphic type argument tuples is \(\min(\max(\PolyMult\cdot n, \PolyFloor),\allowbreak \PolyCap)\), where \(n\) is the number of non-monomorphic type argument tuples associated with the current formula. An important difference with the monomorphic case is that \(n\) depends on the current formula being processed whilst \(m\) does not.
Both in the monomorphic and in the non-monomorphic case, the maximum number of newly generated type argument tuples is fixed per formula and per iteration.

@@

The monomorphisation algorithm is parametrised by many bounds. The first part of the evaluation process is to find appropriate values for these bounds. The second part compares the performance of Zipperposition with and without E on polymorphic problems. The third part compares the performance of different provers on polymorphic problems and their monomorphised counterparts.
Our raw evaluation data is available online.%
\footnote{\url{https://nekoka-project.github.io/pubs/monomorphization_data.zip}}

The benchmarks are taken from version 8.2.0 of the TPTP library~\cite{tptp}. The library contains 1765 problems in the TF1 and TH1 grammars, corresponding respectively to first-order and higher-order logic with rank-1 polymorphism. Because Zipperposition does not support reasoning with real numbers, we removed all problems that include them. In total, our benchmark suite contains 1534 polymorphic problems. We chose as a measure of success for a given prover (or prover configuration) the number of problems that could each be solved by the prover in at most 30 seconds with a single thread.
%This corresponds to the default time limit in Sledgehammer.

@@

We conducted additional informal evaluations on the 500 problem suite to find appropriate default values and test ranges for the monomorphisation bounds. We started the option evaluation process with the base configuration and the following default values for monomorphisation bounds and parameters for E:
%
\begin{minipage}[t]{.5\textwidth}
\vskip0\topsep

\begin{itemize}
   \item \PolyCap: \(\infty\)
   \item \PolyMult: \(1\)
   \item \PolyFloor: \(50\)
   \item substitution ordering: separation
   \item \SubstLimit: \(\infty\)
   \item \MonoSubstsLimit: \(10\)
\end{itemize}

\vskip0\topsep
\end{minipage}\begin{minipage}[t]{.5\textwidth}
\vskip0\topsep

\begin{itemize}
   \item \Limit: \(2000\)
   \item new formulae limit multiplier: \(0\)
   \item monomorphisation timeout: \(20\)
   \item \Loop: \(4\)
   \item E timeout: \(30\)
   \item E call point: \(0\)
\end{itemize}

\vskip0\topsep
\end{minipage}
%
The initial values for the \MonoCap, \MonoMult and \MonoFloor options are irrelevant because the options' values are set when computing Table~\ref{mono_ty_args}.% and used for subsequent tables.

@@

\looseness=-1
Table~\ref{nmon_ty_args} plays the same role as Table~\ref{mono_ty_args} except that it evaluates the bounds limiting the number of new \emph{non-}monomorphic type arguments. The bound values are lower because we found non-monomorphic type arguments to be explosive in preliminary evaluations. The table confirms that non-monomorphic type argument generation drives the creation of useful non-monomorphic formulae. This is indicated by the very low number of problems solved when no new non-monomorphic type arguments are allowed. Performance of the monomorphisation algorithm seems to plateau for some ranges of values and drops off beyond.
%TODO check that 125 is what we get if we just run Zipperposition for 10 seconds (maybe with all additional problems that are easily monomorphised)

@@

%The substitution generation phase occurs once all type arguments have been generated. The bounds limiting the number of new substitutions and the number of monomorphising substitutions per formula are directly related through the role they play in the construction of monomorphising substitutions. The substitution ordering heuristic has a significant impact on monomorphising substitution generation. The ``age'' ordering of substitution orders substitutions generated in earlier iterations are ordered first. The ``random'' ordering randomly shuffles substitutions. Finally, the ``separation'' ordering separates the substitutions into different sets depending on which iteration they were generated in and generates monomorphising substitutions from each of these sets independently. Table \ref{subst_gen_table} shows that the values of bounds limiting the number of substitutions generated has very little impact on overall performance contrary to the ordering heuristic.
The substitution generation phase occurs once all type arguments have been generated. The bound limiting the number of monomorphising substitutions per formula is directly related to the ordering which dictates how such monomorphising substitutions are generated. The substitution ordering heuristic has a substantial impact on monomorphising substitution generation. The ``age'' ordering of substitution orders substitutions generated in earlier iterations are ordered first. The ``random'' ordering randomly shuffles substitutions. Finally, the ``separation'' ordering separates the substitutions into groups of substitutions generated in the same iteration and generates monomorphising substitutions from each of these groups independently. Table \ref{subst_gen_table} shows that the values of bounds limiting the number of monomorphising substitutions only seems to affect performance when the ``separation'' heuristic is used.
% TODO find a better term than "separation heuristic" maybe separation ?

@@

For larger problems, the monomorphisation algorithm may time out despite the bounds. In these cases neither Zipperposition nor E will have had a chance to try to solve the problem. To avoid this, a timer can interrupt the monomorphisation algorithm, after which Zipperposition resumes normal operation. Table~\ref{mono_time} tests the amount of time that monomorphisation is allowed to run for against the number of iteration of the monomorphisation algorithm. Neither parameter seems to have much of an effect on the performance of the algorithm.

@@

\centering\begin{tabular}{@{}l*{3}{>{\centering\arraybackslash}p{6em}}@{}}
   \toprule
   & without E & with E & union \\
   \midrule
   500 problem suite   & 160 & 198 & 207 \\ % -9 +47
   1034 problem suite & 337 & 410 & 434 \\ % -24 +97
   \bottomrule
\end{tabular}
\label{zipp_eval}
\end{table}

@@

\begin{table}[t!]
\caption{Evaluation of native polymorphism vs.\ monomorphisation}

@@

To evaluate the usefulness of iterative monomorphisation as an alternative to native polymorphism, two competitive higher-order polymorphic provers were run on the 1034 problem suite. We chose Leo-III and Zipperposition. We unfortunately had to exclude Vampire because of parsing issues.
